### 2024-05-29 journal/response.txt

Some explanation has been added in the text (line 526) [*** check line #].

> line 620 (line 46 in the code). Suggestion: consider to put the
> check whether the new block B contains no operations
> (num_deqs+num_enqs==0) immediately after the calculation of num_enq
> and num_deqs. First, complete the calculations that depend on the
> values of num_enq and num_deq, and then advance to the check
> whether v is root. If the change will be done, change the order of
> the explanation in lines 715 - 723, right column, accordingly.

Done.

> line 634 (line 58 of the code). In the comment, probably add an
> additional remainder that D(v.blocks[b]) and D(root.blocks[b'])
> represent the sequence of the dequeues represented by the
> corresponding blocks.

Added reminder in caption of Figure.

> line 640 (line 64 of the code). Add a comment that gives some
> intuition for this line.

> lines 643 - 645 (lines 67 - 69 of the code). Add comments that help
> to understand the calculation of the dequeue's rank.

> line 658 (line 81 in the code). Add additional comment that
> doubling search is used here to find the range of the binary
> search.

Done.  Some text has been added to the text description of FindResponse to explain that the doubling search first finds the left end of the range and then performs a binary search.

> line 696, right column. "... and calls Refresh on $v$ twice in
> lines 226 and 227". The lines numbers should be 17 and 18.
>
> line 701, right column. "... prior to line 226" should be "...
> prior to line 17"

These have been fixed.

> line 706, right column. Add a description of the code in lines 26 -
> 29 of Refresh.

> lines 771 - 790, left column. Add explicit references to the lines
> of the code in the description of GetEnqueue().

This has been done, and the description has been expanded.

> line 828, right column. Probably emphasize: if B and B' are two
> *different* blocks at the same depth...

> line 870, right column, Probably should be "By the definition of
> *direct* subblocks (3.3)...".

Yes, clarified.

> lines 887, 892. left column. It look like there are two unnecessary
> empty lines.

This seems to be a strange effect of the double-spacing option for the draft.

> line 919, right column. Should it be "the value computed in line
> *42*" (instead of 41) ?

Fixed.

> line 983, right column. "... close to the time ..." is not clear in
> the context of asynchronous execution. Probably state it
> differently: "We do this by showing that B.super is written after B
> is installed, but before B_s is installed."

This has been restated.

> line 1003, right column, Probably should be "By the definition of
> *direct* subblocks (3.3)...".

Fixed.

> line 1102, right column. A more detailed explanation would be
> helpful. Particulary, if op_1 precedes op_2, in which blocks op_1
> and op_2 can appear so that by (3.2) op_1 precedes op_2 in L ?

This has been expanded.

### 2024-05-29 journal/algorithm.tex
working on comments from reviewer 1

Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations within the same block, since they must be concurrent.

@@

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{\# of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
	\end{itemize}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{Object} \fld{element}
  		\Comment{\var{x} for \opa{Enqueue}{x} operation; otherwise \nl}
	\end{itemize}
	\item[\com] Blocks in the root \node\ have the following additional field
	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{int} \size%
  		\Comment{size of queue after performing all operations up to end of this block}	\end{itemize}
\end{itemize}

%\Statex {\com\  Blocks in internal nodes have the following additional fields}
%\Statex $\blacktriangleright$ \typ{InternalBlock} extends \typ{Block} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
%\begin{itemize}
%	\item \typ{int} \eleft, \eright
%  		\Comment{index of last direct subblock in the left and right child}
%\end{itemize}

%\Statex {\com\ Blocks in leaf nodes have the following additional field}
%\Statex $\blacktriangleright$ \typ{LeafBlock} extends \typ{Block} \sf{\com\ the following additional field is used only for blocks in leaves}
%\begin{itemize}
%  \item \typ{Object} \var{element}
%  \Comment{if the block's operation is enqueue(x) then element=x, otherwise element=null.}
%\end{itemize}

%\Statex {\com\ Blocks in the root node have the following additional field}
%\Statex $\blacktriangleright$ \typ{RootBlock} extends \typ{InternalBlock} \sf{\com\ the following additional field is used only for blocks in the root}
%\begin{itemize}
%  \item \typ{int} \size%
%  \Comment{size of the queue after performing all operations up to the end of this block}
%\end{itemize}

\end{algorithmic}
\caption{Objects used in the \ordering\ tree data structure.\label{object-fields}}
\end{figure*}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else \Comment{First, find the superblock of $v.\fld{blocks}[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\label{idsup1}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
	    \EndIf\label{idsup3}
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \Comment{queue is empty when dequeue occurs}\label{returnNull}
    \Else \ \Comment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
		\State use binary search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}
\end{algorithmic}
\caption{Queue implementation's helper routines. \label{code2}}
\end{figure*}

@@

    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
        \Else \Comment{required enqueue is in \var{v}.\fld{right}}
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\
        	 that $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}

@@

An \opemph{Enqueue}(\var{e}) 
\rev{creates a new Block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
that the Block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.}

A \opemph{Dequeue} also 
\rev{creates a new Block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
that the Block represents a \op{Dequeue}.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

@@

A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
in \var{v}'s
children and tries to append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
Otherwise, the CAS at line \ref{cas} tries to install $B$ into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a  block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
from \var{h} to $\var{h}+1$
and fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
A \op{Refresh} may pause after a successful CAS before calling \op{Advance} at line \ref{advance},
so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.

@@

To compute the response of the $i$th \op{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If not, line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
the \var{e}th enqueue.
Since the enqueue is linearized before the dequeue, $b_e\leq b$.  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} traces down through the tree to find the required enqueue in a leaf.

\opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
$i$th enqueue in the $b$th block $B$ of \typ{Node} $\var{v}$. 
It recursively finds the location of the enqueue in each node along the path from $v$ to a leaf, which stores the argument explicitly.
\op{GetEnqueue} first determines which child of \var{v} contains the enqueue, and
then finds the range of blocks within that child that are subblocks of $B$ using information stored
in $B$ and the block that precedes $B$ in $v$.
\op{GetEnqueue} finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
field (line~\ref{getChild}) and proceeds recursively down the tree.

>>>>

Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's \fld{blocks} array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations within the same block, since they must be concurrent.

@@

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{\# of enqueues, dequeues in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
	\end{itemize}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{Object} \fld{element}
  		\Comment{\var{x} for \opa{Enqueue}{x} operation; otherwise \nl}
	\end{itemize}
	\item[\com] Blocks in the root \node\ have the following additional field
	\begin{itemize}[leftmargin=0mm]
		\item[$\bullet$] \typ{int} \size%
  		\Comment{size of queue after performing all operations up to end of this block}	\end{itemize}
\end{itemize}
\end{algorithmic}
\caption{Objects used in the \ordering\ tree data structure.\label{object-fields}}
\end{figure*}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}} 
    \State \linecomment{return $\langle\var{b}', \var{i}'\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is $(\var{i}')$th dequeue of $D(\var{root}.\fld{blocks}[\var{b}'])$}
    \State \linecomment{Precondition: $\var{v}.\fld{blocks}[\var{b}]$ is not \nl, was propagated to root, and contains at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCase}
    \Else \Comment{first, find the superblock of $v.\fld{blocks}[b]$}
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\State \var{sup} \assign\ \var{v}.\fld{blocks}[\var{b}].\fld{super}\Comment{\rev{actual index of superblock is either \var{sup} or $\var{sup}+1$}}\label{idsup1}
	    \If{$\var{b} > \var{v}.\fld{parent.blocks}[\var{sup}].\fld{end\sub{dir}}$} \var{sup} \assign\ $\var{sup}+1$\label{supertest}\label{idsup2}
	    \EndIf\label{idsup3}
	    \State \linecomment{compute index \var{i} of dequeue in superblock}
	    \State \linecomment{\rev{add \# dequeues in blocks of $v$ before $v.\fld{blocks}[b]$ within superblock $v.\fld{parent.blocks}[\var{sup}]$}}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ 
	    		$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\edir].\fld{sum\sub{deq}}$}\label{computeISuperStart}
        \If{$\fld{dir} = \fld{right}$} \Comment{\rev{add \# dequeues from left sibling}}
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}].\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\var{v}.\fld{parent.blocks}[\var{sup}-1].\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRight}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, \var{sup}, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \com\ find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \Comment{queue is empty when dequeue occurs}\label{returnNull}
    \Else \ \Comment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ $\var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size$ \label{computeE}
		\State use \rev{doubling} search to find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\Comment{find rank of enqueue within its block}\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}
\end{algorithmic}
\caption{Queue implementation's helper routines. \rev{Recall that $D(B)$ is the sequence of dequeues represented by block $B$.}\label{code2}}
\end{figure*}

@@

    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from \var{v}.\fld{left}}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{left}}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from \var{v}.\fld{right}}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \fld{dir} \assign\ \fld{left}\label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
        \Else \Comment{required enqueue is in \var{v}.\fld{right}}
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$ \Comment{\rev{deduct \# enqueues in $v.\fld{blocks}[b]$ from $v.\fld{left}$}}
        \EndIf \label{endChooseDir}
        \State \linecomment{find enqueue's block in \var{v}.\fld{dir.blocks} and its rank within block}
        \State \hangbox{use binary search to find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[$\var{b}-1$].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] such\\
        	 that $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}

@@

An \opemph{Enqueue}(\var{e}) 
\rev{creates a new block at line \ref{enqNew} and sets its \fld{sum} fields to indicate
that the block represents an \op{Enqueue}.  The enqueued item \var{e} is stored in the block's \fld{element}
field.
The call to \op{Append}, described below, on line \ref{enqAppend}
appends the \block\ to the process's leaf and ensures it is propagated to the root.}

A \opemph{Dequeue} also 
\rev{creates a new block at line \ref{deqNew} and sets its \fld{sum} fields to indicate
that the block represents a \op{Dequeue}.
It then calls \op{Append} at line \ref{deqAppend} to append the \block\ to the process's leaf and ensure
it is propagated to the root.
Then,} the \op{Dequeue} calls \op{IndexDequeue} to compute
its position in the root 
 and then calls \op{FindResponse} to compute its response.

@@

A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
in \var{v}'s
children and tries to append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
\rev{Lines \ref{startHelpChild1}--\ref{endHelpChild1} ensure that the \fld{head} values
for \var{v}'s children have been advanced to the index of the latest block stored in their \fld{blocks} arrays.}
Line \ref{invokeCreateBlock} creates the new block $B$ to install in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
Otherwise, the CAS at line \ref{cas} tries to install $B$ into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a  block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
from \var{h} to $\var{h}+1$
and fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
A \op{Refresh} may pause between a successful CAS at line \ref{cas} and calling \op{Advance} at line \ref{advance},
so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.

@@

To compute the response of the $i$th \op{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If not, line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
\rev{The routine next searches for}
the index $b_e$ of the block in \var{root}.\fld{blocks} that contains 
the \var{e}th enqueue.
Since the enqueue is linearized before the dequeue, $b_e\leq b$.  To find the left end of the range \rev{to search for $b_e$, \op{FindResponse} does} a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
\rev{Then, the doubling search performs a binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} within that range to find $b_e$.}
Finally, the call to \op{GetEnqueue} \rev{at line \ref{findAnswer}} traces down through the tree to find the required enqueue in a leaf.

\opemph{GetEnqueue}(\var{v, b, i}) returns the argument of the
$i$th enqueue in the $b$th block $B$ of \typ{Node} $\var{v}$. 
It recursively finds the location of the enqueue in each node along the path from $v$ to a leaf, which stores the argument explicitly.
\rev{The test at line \ref{leftOrRight} determines which child of \var{v} contains
the enqueue by checking whether $i$ is less than or equal to the number of  
enqueues in $B$ that came from \var{v}'s left child.
Line \ref{getChild} determines the range of blocks within that child that are subblocks of $B$} 
using information stored
in $B$ and the block that precedes $B$ in $v$,
\rev{and then}
finds the exact subblock containing the enqueue using a binary search on the \fld{sum\sub{enq}}
field \rev{of the child's \var{blocks} array.
Line \ref{getChildIndex} computes the index of the enqueue within the block
in the child, and line \ref{getRecurse}
proceeds recursively down the tree.}
